---
title: "Introduction to R Markdown"
author: "B. Sosnovski"
date: "09/25/2021"
output: 
  html_document: 
    theme: cerulean
---

```{r setup, include=FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

### Markdown

Markdown is a way of adding to a plain text document formatting such as headers, lists, emphasis, quotes, links to URLs and images, and more. 

Markdown allows you to write using an easy-to-read, easy-to-write plain text format, then convert it to HTML.

### R Markdown

R Markdown allows you to write a  HTML, PDF, and/or MS Word documents, a presentation or even a book by just using Markdown syntax and embedded programming code.

R Markdown files have extension **.Rmd** 

The "**R**" in R Markdown stands for the [R statistical computing](https://www.R-project.org/) language.


# Load Packages

```{r}
library(knitr)
# The knitr package extends the basic markdown syntax to include chunks of executable R code
# Cocalc already have this package loaded for you but if you are running your code, say in R Studio, you would need to include the knitr in the library of your document.
```

# Formatting text

I hope you already noticed some of the formatted text above in this rmd file.

Here some basics formatting of text in Markdown.

#### Comments

Note that you can add comments in Markdown documents as follows: ``<!-- your comment -->``. The comments will not be displayed in any output format.

<span style='color:#d2691e'>Compare this line here in the rmd file on the left and the output format in HTML on the right (Note that the comment below is missing!!.</span>

<!--  This will not be displayed in the output format -->

#### Plain Code Blocks

Plain code blocks are displayed in a fixed-width font but not evaluated by R.

```
This text is displayed verbatim
```
You will see this throughout this rmd file to display examples of codes that are not evaluated.


#### Headers
   
   You can include section headers by using pound signs.
   

```
# First-level header

## Second-level header

### Third-level header
```

#### Emphasis


```
*italic*   **bold**

_italic_   __bold__

```
Note that the word *Statistics* is in italic here and the work **markdown** is in bold.
  
#### Lists

###### Unordered list

* Item 1
* Item 2
    + Item 2a
    + Item 2b

The code for the list above is:

```
* Item 1
* Item 2
    + Item 2a
    + Item 2b
    
```

###### Ordered list

1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b


The code for the list above is:

```
1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b
```
##### Hyperlinks

To add hyperlinks we surround the links with brackets, and then provide the link address in parentheses, like this [QCC](https://www.qcc.cuny.edu).

```
[QCC](https://www.qcc.cuny.edu)
```

# Embedding R code


You can embed an R code chunk like this:

```{r}
summary(cars)
# cars is a built-in data set in R
```

Note that the line above is a comment in the R code, and that you can add comments in the R code chunk with a pound sign. Another example:

```{r}
2*3
# This is purely a comment that will be ignored by the R evaluator.
```

To add an inline code use a pair of backticks and the letter r. For example, `r 2*3`. We can also mark text as inline code by using just the pair of backticks. For example, `code`.

You can also embed plots in your rmd files. For example:

```{r, echo=FALSE}
plot(cars)
```

You can do a lot of things in a code chunk: produce text output, tables, or graphics. You have fine control over all these output via chunk options, which can be provided inside the curly braces (between `{r and }`).

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.



# Writing mathematical equations and formulas

Of course, you can include math formulas by using dollar signs around the equation in Latex syntax. 

Here is the code for writing the equation for the area of a circle $A= \pi r^2$.

##### In line

```
$ A= \pi r^2 $
``` 

##### In equation display

```
$$ A= \pi r^2 $$
```


# References

Parts of the material used in this document are adaptations or excerpt from the following:

* Grolemund, Garrett. "Introduction to R Markdown."  July 16, 2014. https://rmarkdown.rstudio.com/articles_intro.html

* "Markdown Basics." https://rmarkdown.rstudio.com/authoring_basics.html

